Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance

### 2013-11-28 HRUIofMovingObjects/jrnl_article.tex
Moving objects article: Done some writing and added updated figures.

\begin{abstract}
%\boldmath
If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be local shift-invariant. This can for instance happen if the axial and lateral sampling is insufficient. The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images. Increased lateral resolution should demand denser lateral sampling. In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.

In this paper we investigate the local lateral shift-invariance of the Capon beamformer when imaging moving objects. We then discuss and investigate different methods in order to improve it. We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally lateral shift-variant. We then show that the locally lateral shift-invariance can be improved by simple phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ rad). Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied. The figure shows that when the DAS beamformer is used with a beamspacing equal to the system resolution, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.

\begin{figure}[!t]
\centerline{
\includegraphics[width=\linewidth]{\img steerd_respons_capon_das_resolution.eps}
}
\caption{Comparison of drop in output power in between receive beams. Half the system resolution is plotted to show where two beams with a spacing equal to the system resolution would intersect. The steered response of both delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$) is plotted. A monochromatic plane wave is located at zero degrees.}
\label{fig:das_capon_beams}
\end{figure}

@@

In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate. In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data. Finally we give our conclusion in Section \ref{sec:con}.

@@

The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}

Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}. From (\ref{eq:capon_weights}) it is clear that complex data can lead to complex weights. The weights can therefore have an asymmetric K-space response which basically means that the main lobe and zeros does not have to be symmetric around the origin. The beam can therefore be micro-steered by phase rotation.

@@

The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally lateral shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer. In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the loss in between beams is smaller than 1 dB. However, depending on light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer. For each lateral position of the point scatterer the root-mean-square (RMS) is calculated in the range dimension, and the resulting beam profiles are stacked to produce an image. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot. The system is said to be locally lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus a straight line ($f(x)=x$). We name this measure P-MAE. The MAE of the amplitudes for the same set of maximum points versus the maximum in the whole plot is also measured. We call this measure A-MAE. The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal. The maximum amplitude deviation along the max-point trace, or the scalloping loss, is also calculated and reported. We present all these numbers together with the LSV-plots.

Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}). The red dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup mimicking the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and x cycles long pulse \comment{Add correct number of cycles}. We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant. The P-MAE and A-MAE are also minimal.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
}
\caption{LSV-plots of delay-and-sum beamforming. If the contours are diagonal the imaging system is said to be locally lateral shift-invariant. Red vertical lines are plotted where transmit-receive beam pairs are located. a) Sampling corresponding to (\ref{eq:resolution}). The system is locally lateral shift-invariant. b) Undersampling with a factor of two. The system is no longer locally lateral shift-invariant.}
\label{fig:das}
\end{figure*}

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon1}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}\label{fig:capon2}
}
}
\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}). a) Capon beamforming with $d=1/100$ as diagonal loading. The system is not locally lateral shift-invariant. b) The diagonal loading factor $d$ is increased to 1. The system is still not locally lateral shift-invariant.}
\label{fig:capon}
\end{figure*}

@@

\subsection{Oversampling by Parallel Receive Beamforming}
Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam. Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g. cardiac ultrasound imaging. However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}. In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming. In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording. In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality, and the LSV-plots for PRB is therefore not included.

>>>>

\begin{abstract}
%\boldmath
If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be local shift-invariant. This can for instance happen if the axial or lateral sampling is insufficient. The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images. Increased lateral resolution should demand denser lateral sampling. In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.

In this paper we investigate the local lateral shift-invariance of the Capon beamformer when imaging moving objects. We then discuss and investigate different methods in order to improve on it. We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally lateral shift-variant. We then show that the locally lateral shift-invariance can be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ radians). Impinging on the array is a monochromatic plane wave originating from a source moving parallel to the array surface. The plot shows the steered response of the source when it is located at plus and minus half the system resolution. No spatially white noise present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied. The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, will yield a 3.92 dB scalloping loss \cite{Harris1978} if the source is moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.

\begin{figure}[!t]
\centerline{
\includegraphics[width=\linewidth]{\img fig1.eps}
}
\caption{Comparison of drop in output power in between receive beams. The steered response of both delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$) is plotted for two different time instances. A monochromatic plane wave is located at an angle equal to half the system resolution, and is moved a distance corresponding to the system resolution for the second steered response.}
\label{fig:das_capon_beams}
\end{figure}

@@

In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout. In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} data. Finally we give our conclusions in Section \ref{sec:con}.

@@

The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}. From (\ref{eq:capon_weights}) it is clear that complex data can lead to complex weights. The weights can therefore have an asymmetric K-space response which basically means that the main lobe and zeros does not have to be symmetric around the origin. The beam can therefore be micro-steered by phase rotation.

@@

The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally lateral shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer. In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the loss in between beams is smaller than 1 dB. However, depending on viewing light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer. For each lateral position of the point scatterer the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked to produce an image. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot. The system is said to be locally lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus a straight line ($f(x)=x$). We name this measure P-MAE. The MAE of the amplitudes for the same set of maximum points versus the maximum in the whole plot is also measured. We call this measure A-MAE. The first measure quantifies the amount of geometric distortion, and the latter quantifies the peak gain variation or in our case the scalloping loss. The maximum amplitude deviation along the maximum-point trace, or the maximum scalloping loss, is also calculated and reported. We present all these numbers together with the LSV-plots.

Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}). The red dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup mimicking the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. We see that the contours are approximately diagonal and therefore we conclude that the system is locally lateral shift-invariant. The P-MAE and A-MAE are also minimal.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
}
\caption{LSV-plots of delay-and-sum beamforming. If the contours are diagonal the imaging system is said to be locally lateral shift-invariant. Red vertical lines are plotted where transmit-receive beam pairs are located. a) Sampling corresponding to (\ref{eq:resolution}). The system is locally lateral shift-invariant. b) Undersampling with a factor of two. The system is no longer locally lateral shift-invariant.}
\label{fig:das}
\end{figure*}

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig3a.eps}\label{fig:capon1}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig3b.eps}\label{fig:capon2}
}
}
\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}). a) Capon beamforming with $d=1/100$ as diagonal loading. The system is not locally lateral shift-invariant. b) The diagonal loading factor $d$ is increased to 1. The system is still not locally lateral shift-invariant.}
\label{fig:capon}
\end{figure*}

@@

\subsection{Oversampling by Parallel Receive Beamforming}
Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam. Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g. cardiac ultrasound imaging. However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007, Denarie2011, Denarie2013}. In \cite{Rabinovich2013} it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming. In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording. In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality, and the LSV-plots for PRB is therefore not included.
